Natural variations in water discharge quantity have an inconsistent impact on sediment transport capacity in subglacial channels

### 2023-10-05 Delaneyetal2023b.tex
more small bits

\title{Natural variations in water discharge quantity has an inconsistent impact on sediment transport capacity in subglacial channels}

@@

Second, to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies outside of the range between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies outside of the range $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
Variability of an output is represented by the  standard deviation of an output over the course of the season.

Lastly, outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation (i.e. correlation depends on the ordering of values from the outputs and the observations).
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Outputs and observations at $15$\,\unit{min} intervals are aggregated across timescales of up to $15$\,\unit{days} by averaging values one half of the timescale before and after a given point time.

@@

\subsection{Role of water discharge in sediment transport capacity}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Parameterization outputs resulting from the hydrographs shown in panels a and e for \alpine cases (a-d) and \icesheet{} (e-h).
    Gray (black) lines in a and e represent the subglacial channel size (water discharge).
    In other panels, blue lines represent outputs from the subglacial channel, while orange lines represent the subaerial channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows denote some events where  variables peak in the subglacial channel prior to the subaerial one.
    Insets in e-h show the  peak melt event \icesheet{} denoted by the shaded area.
    Note that the x and y axes are different for \alpine{} and \icesheet{}.
      }
  \label{fig:model_outs}
\end{figure}

@@

Over the course of the season, large peaks in water velocity and shear stress, along with width-integrated shear stress, occur in response to water discharge variations in both \alpine{} and \icesheet{}.
In the subglacial channel, peaks generally occur over the period when water discharge increases at the fastest rate (Figure~\ref{fig:model_outs}).
Alternatively, in the subaerial channel, peaks in these sediment transport variables occur when water discharge is highest.
The timing of the greatest rate of change in water discharge occurs before the peaks in water discharge.
Furthermore, the subglacial outputs are generally larger than the subaerial values, thus more efficient at transporting sediment.
Here, sediment could be deposited in the proglacial area as the water depressurizes, then remobilized when sediment transport capacity peaks in the subaerial system.

@@

Each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress, creating a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari} a--f).
In the subglacial channel, however, the persistence of high water velocity, shear stress, and width-integrated shear stress across a range of discharges suggests that substantial sediment transport may occur across much of the season (Figure~\ref{fig:Qw_vari} a--f).
For instance, very high  water velocity values and shear stresses can occur across minimal water discharge at $2.5$\,\unit{m}$^3$\,\unit{s}$^{-1}$ to the maximum water discharge at over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
In \icesheet, mean values of water velocity can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
When considering the subglacial channel's evolving width, width-integrated shear stress  across the channel generally increases with water discharge, with improved rank correlation compared to water velocity or shear stress (Figure~\ref{fig:Qw_vari} a--f).
Yet even the width-integrated shear stress  can vary substantially, with the highest values occurring at water discharge values ranging from roughly $11$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ to over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet case, with reduced variability in water discharge (Figure~\ref{fig:Qw_vari} c, f).

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\linewidth]{Fig4.png}
    \caption{Variability (standard deviation) in water velocity, shear stress, and width integrated shear stress for different time aggregations across the range of subglacial and subaerial channel shapes and friction factors.
      Shaded areas denote the minimum and maximum of the $100$ parameter combinations tested (Section~\ref{sect:imp}).
      Solid lines denote  mean values.
      %Top row shows results from the \alpine{} case, while bottom row show outputs from the \icesheet{} case
      Time intervals shown with markers denote $15$\,\unit{min}, $1$\,\unit{hr}, $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d} aggregation periods.
      Text in columns shows rank correlation between the variable and sediment discharge for the corresponding time period for the subglacial (right column) and subaerial (left column) channels.
    }
    \label{fig:multi_run}
  \end{figure}

The second application examines the impact of a variety of channel shapes and friction factors on the relationship between water discharge and sediment transport capacity.
Across the range of parameter values examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one.
In \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge (Figure~\ref{fig:multi_run}).
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

It is prescient to evaluate the time scales over which the subglacial channels may respond to changing water discharge to understand the influence of variability in sediment transport capacity.
In both cases, the variability in velocity, shear stress, and width-integrated shear stress decrease after $1-5$ day aggregations (Figure~\ref{fig:multi_run}).
Yet, substantial differences in variability persist in periods of up to $15$ days, as the variability in aggregations periods longer than this might be conflated with seasonal variations in water discharge.
Note that lags and time of peak sediment transport conditions still occur over these periods (Figure~\ref{fig:model_outs}), while the variability in sediment transport parameters decreases with larger time aggregation (Figure~\ref{fig:multi_run}).

@@

Quantities that control sediment transport capacity, such as water velocity, shear stress, and width-integrated shear stress, vary differently in response to water discharge in subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}).
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}; Table~\ref{tab:Qs}).
Note that sediment transport relationships describe sediment flux per unit width, so that the integrated capacity across channel occurs after application of the exponent, unlike the width-integrated shear stress evaluation here \cite<Equations\,\ref{eq:tautg} and \ref{eq:tautf}; c.f.>{tucker1997}.
%Note also that both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress is homogeneous across the channel bed, despite the limitations of this assumption \cite<e.g.>{yager2018}.

% The rapid increases and decreases in subglacial sediment transport capacity compounds the variability in sediment transport already present in subaerial systems \cite<Figure~\ref{fig:Qw_vari}; >{williams1989,jerolmack2010}.
Sediment transport capacity in subglacial channels rapidly responds to changes in water discharge compared to their subaerial counter parts \cite{alley1997}.
In subglacial R\"othlisberger channels where water discharge evolves more slowly than channel size or subaerial channels the width-integrated shear stress response to water discharge is $ \tau_{tg}\, \propto\,  Q^{\sim\frac{4}{5}}$ (Table\,\ref{tab:eqs1}).
However, subglacial channels rarely operate in a steady state \cite<e.g.>{gimbert2016}.
For sufficiently short time periods, channel width and wetted area are fixed in the subglacial system (assuming pipe flow) so that the response of width-integrated shear stress is $ \tau_{tg}\, \propto\,  Q^{2}$ (Table\,\ref{tab:eqs1}).
Appendix~\ref{sect:scaling} provides a complete description of sediment transport capacity in different drainage conditions.

The strong correlation between hydraulics and water discharge in \icesheet{} could result from being closer to an equilibrium between variations in the channels' wetter area and water discharge with smaller diurnal variations in amplitude (Figures~\ref{fig:model_outs} and ~\ref{fig:Qw_vari}).
Conversely, the correlation between hydraulics and water discharge in \alpine is smaller due to the larger discharge variations. 
In both cases, hydraulics could trend toward equilibrium by averaging the discharge over $1-5$ days, when the variability in subglacial channels approaches that of subaerial one (Figure~\ref{fig:multi_run}).

In subglacial channels, peaks in water velocity, shear stress and width-integrated shear stress occur with the fastest rate of water discharge increase relative to the channel's growth (Figure~\ref{fig:model_outs}).
In turn, subglacial channels grow with increasing water discharge, effectively reduces sediment transport capacity with the stabilization of water discharge at its peak.
Similar feedback occurs in subaerial systems \cite{phillips2016}, albeit at timescales longer than days \cite{gimbert2016}.

@@

First, these measurement stations lie $0.5$ to $2$\,\unit{km} downstream of the glacier terminus \cite{cowton2012,felix2022}, a long enough distance for the pressurized subglacial regime to transition to the open channel subaerial one.
Actually, the transition of subglacial to subaerial conditions may occur underneath glacier terminus \cite{perolo2018}, especially if closure rates of the subglacial channels are slow  \cite{rothlisberger1972}.
Furthermore, if a transport-limited regime persists at the glacier margin, then sediment can be easily mobilized, and subaerial processes could be represented just a short distance downstream.

>>>>

\title{Natural variations in water discharge quantity have an inconsistent impact on sediment transport capacity in subglacial channels}

@@

Second, to further characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies between $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
Variability of an output is represented by the  standard deviation of an output over the course of the season.

Lastly, outputs are compared with the sediment discharge records of the glaciers using Spearman rank correlation, dependent on the ordering of values.
Rank correlation reduces the impact of the non-linear response of sediment transport capacity to the hydrological forcing.
Outputs and observations at $15$\,\unit{min} intervals are aggregated across timescales of up to $15$\,\unit{days} by averaging values one half of the timescale before and after a point time.

@@

\subsection{Role of water discharge in sediment transport capacity}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Parameterization outputs from the hydrographs n panels a and e for \alpine cases (a-d) and \icesheet{} (e-h).
    Gray (black) lines in a and e represent the subglacial channel size (water discharge).
    Blue lines represent outputs from the subglacial channel, while orange lines represent the subaerial channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows denote where variables peaking in the subglacial channel prior to the subaerial one.
    Insets in e-h show the  peak melt event \icesheet{} denoted by the shaded area.
    %Note that the x and y axes are different for \alpine{} and \icesheet{}.
  }
  \label{fig:model_outs}
\end{figure}

@@

In the subglacial channel, peaks generally occur when water discharge increases at the fastest rate (Figure~\ref{fig:model_outs}).
Alternatively, in the subaerial channel,  sediment transport variables peak  hen water discharge is highest.
The timing of the greatest rate of change in water discharge occurs before the peaks in water discharge.
Furthermore, the subglacial outputs are generally larger than the subaerial values, thus more efficient at transporting sediment.
Here, sediment could be deposited in the proglacial area as the water depressurizes, then remobilized when sediment transport capacity peaks in the subaerial system.

@@

Each water discharge value in the subaerial channel results in a unique water velocity, shear stress, and width-integrated shear stress, creating a perfect rank correlation between the variables (Figure~\ref{fig:Qw_vari} a--f).

In the subglacial channel, however, the persistence of high water velocity, shear stress, and width-integrated shear stress across a range of discharges suggests that substantial sediment transport may occur across  (Figure~\ref{fig:Qw_vari} a--f).

For instance, very high  water velocity values and shear stresses can occur across minimal water discharge at $2.5$\,\unit{m}$^3$\,\unit{s}$^{-1}$ to the maximum water discharge at over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
In \icesheet, mean values of water velocity can occur at water discharges between roughly $150$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ and $310$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
When considering the subglacial channel's evolving width, width-integrated shear stress  across the channel generally increases with water discharge, with improved rank correlation compared to water velocity or shear stress (Figure~\ref{fig:Qw_vari} a--f).
Yet even the width-integrated shear stress  can vary substantially, with the highest values occurring at water discharge values ranging from roughly $11$ \,\unit{m}$^3$\,\unit{s}$^{-1}$ to over $17$ \,\unit{m}$^3$\,\unit{s}$^{-1}$.
The variability in width-integrated shear stress is less pronounced in the \icesheet case, with reduced variability in water discharge (Figure~\ref{fig:Qw_vari} c, f).

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\linewidth]{Fig4.png}
    \caption{Variability (standard deviation) of output variables for different time aggregations across the range of subglacial and subaerial channel shapes and friction factors.
      Shaded areas denote the range of the $100$ parameter combinations tested (Section~\ref{sect:imp}).
      Solid lines denote  mean values.
      %Top row shows results from the \alpine{} case, while bottom row show outputs from the \icesheet{} case
      Time intervals shown with markers denote $15$\,\unit{min}, $1$\,\unit{hr}, $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d} aggregation periods.
      Text in columns shows rank correlation between the variable and sediment discharge for the corresponding time period for the subglacial (right column) and subaerial (left column) channels.
    }
    \label{fig:multi_run}
  \end{figure}

Across the range of channel shape parameters and friction factors  examined, variability in water velocity and shear stress remains higher in the subglacial system compared to the subaerial one.
In \alpine{}, the variability in width-integrated shear stress is comparable between the two systems, due to the variable water discharge (Figure~\ref{fig:multi_run}).
However, in \icesheet{}, with lower water discharge fluctuations, the variability in width-integrated shear stress remains higher in the subglacial system across the time aggregation periods.

It is prescient to evaluate the time scales over which the subglacial channels may respond to changing water discharge to understand the influence of variability in sediment transport capacity.
In both cases, the variability in velocity, shear stress, and width-integrated shear stress decreases after $1-5$ day aggregations (Figure~\ref{fig:multi_run}).
Yet, substantial differences in variability persist in periods of up to $15$ days.
Aggregations periods longer than this might be conflated with seasonal variations in water discharge.
Note that lags and time of peak sediment transport conditions still occur over these periods (Figure~\ref{fig:model_outs}), while the variability in sediment transport parameters decreases with larger time aggregation (Figure~\ref{fig:multi_run}).

@@

Quantities that control sediment transport capacity (water velocity, shear stress, and width-integrated shear stress) vary differently in response to water discharge in subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}).
In sediment transport relationships that evaluate sediment discharge capacity, such as in \citeA{meyer1948} or \citeA{engelund1967}, shear stress is scaled to the power of $\frac{3}{2}$ or $\frac{5}{2}$, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}; Table~\ref{tab:Qs}).
Note that sediment transport relationships describe sediment flux per unit width, so that the integrated capacity across channel occurs after application of the exponent, unlike the width-integrated shear stress evaluation here \cite<Equations\,\ref{eq:tautg} and \ref{eq:tautf}; c.f.>{tucker1997}.
%Note also that both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress is homogeneous across the channel bed, despite the limitations of this assumption \cite<e.g.>{yager2018}.

% The rapid increases and decreases in subglacial sediment transport capacity compounds the variability in sediment transport already present in subaerial systems \cite<Figure~\ref{fig:Qw_vari}; >{williams1989,jerolmack2010}.
Sediment transport capacity in subglacial channels responds more rapidly to changes in water discharge compared to their subaerial counter parts \cite{alley1997}.
In subglacial R\"othlisberger channels where water discharge evolves more slowly than channel size or subaerial channels the width-integrated shear stress response to water discharge is $ \tau_{tg}\, \propto\,  Q^{\sim\frac{4}{5}}$ (Table\,\ref{tab:eqs1}).
However, subglacial channels rarely operate in a steady state \cite<e.g.>{gimbert2016}.
For sufficiently short time periods, channel width and wetted area are fixed in the subglacial system (assuming pipe flow) so that the response of width-integrated shear stress is $ \tau_{tg}\, \propto\,  Q^{2}$ (Table\,\ref{tab:eqs1}).
Appendix~\ref{sect:scaling} provides a complete description of sediment transport capacity in different drainage conditions.

The stronger correlation between subglacial hydraulics and water discharge in \icesheet{} could result from the channel being closer to equilibrium between variations in the channels' wetter area and water discharge with smaller diurnal variations in amplitude (Figures~\ref{fig:model_outs} and ~\ref{fig:Qw_vari}).
Conversely, the correlation between hydraulics and water discharge in \alpine is smaller due to the larger discharge variations. 
In both cases, hydraulics could trend toward equilibrium by averaging the discharge over $1-5$ days, when the variability in subglacial channels approaches that of subaerial one (Figure~\ref{fig:multi_run}).

In subglacial channels, peaks in water velocity, shear stress and width-integrated shear stress occur with  fast rates of water discharge increase relative to the channel's growth (Figure~\ref{fig:model_outs}).
In turn, subglacial channels grow with increasing water discharge, effectively reduces sediment transport capacity with the stabilization of water discharge at its peak.
Similar feedback occurs in subaerial systems \cite{phillips2016}, albeit at timescales longer than days \cite{gimbert2016}.

@@

First, these measurement stations lie $0.5$ to $2$\,\unit{km} downstream of the glacier terminus \cite{cowton2012,felix2022}, a long enough distance for the pressurized subglacial regime to transition to the open channel subaerial one.
The transition of subglacial to subaerial conditions may occur underneath glacier terminus \cite{perolo2018}, especially if closure rates of the subglacial channels are slow  \cite{rothlisberger1972}.
Furthermore, if a transport-limited regime persists at the glacier margin, then sediment can be easily mobilized, and subaerial processes could be represented just a short distance downstream.
